Dependent Types and Martin-L\"of Type Theory: An Introduction

### 2016-08-09 paper.tex
worked hott part

\section{Introduction}
We present an introduction to dependent types and (Martin-L\"of) type theory.
At its core, type theory seeks to serve as a possible foundation of mathematics,
one that is based on a different logical foundation. This proof-relevant
form of mathematics is the basis in which types can be used to model logical
propositions and is the core idea of type theory. As such, not only does type
theory have a viable application in mathematics but also a growing interest
in the field of computer science where types find themselves in great use in
many popular programming languages. In both fields, type theory provides a
richer world of types and expressibiilty.

The study of type theory goes hand in hand with the study of dependent types,
which is why we first take the time to familiarize ourselves with them
(Section 2). We then continue with the fundamental concepts of type theory of
Per Martin-L\"of \cite{}. For this, we recommend the reader have a running proof
assistant, such as Agda. In our examples, we refer to Agda syntax rules, and
using a proof assistant should help develop the necessary thought processes to
successfully do type theoretic proofs.

@@

\subsection*{Simply-Typed (STLC)}
\subsubsection*{Role: Base System}
The STLC is not a dependently typed system. In fact, the term {\em dependent type}
is drawn in constrast with the term {\em simple type}, where the STLC derives
its namesake. A simple type ({\tt t}) is either a {\em primitive type}, {\tt Prim},
which is the set of all types ``pre-defined'' (e.g. lists, numbers, booleans),
or a {\em function type} between other simple types. Thus, simple types have the
following grammar:
\begin{center}
\begin{minipage}{0.5\textwidth}
\begin{code}
Prim || t → t
\end{code}
\end{minipage}
\end{center}
The terms of the STLC are the terms of the \textit{untyped \(\lambda\)-calculus},
with the exception of terms that use \textit{self-application} (i.e., Y
combinator, Omega, etc.). An example term, annotated with its type, is generally
written as follows:
\begin{center}
\begin{minipage}{0.5\textwidth}
\begin{code}
example : ℕ → ℕ
example = λ (n : ℕ) → n
\end{code}
\end{minipage}
\end{center}
Or, equivalently,
\begin{center}
\begin{minipage}{0.5\textwidth}
\begin{code}
example = λ n → n
\end{code}
\end{minipage}
\end{center}
whenever the type of the argument (viz. {\tt n}) is easily inferable. In this
case, since {\tt example} is of type {\tt ℕ → ℕ}, the argument {\tt n} has
to be of type {\tt ℕ}, therefore leaving its type annotation unnecessary.

The STLC has limited expressibilty but has the benefit of no longer having any
infinite calculations \cite{} and is thus the appropriate foundation for
deriving a type theoretic system. This is because the STLC lends itself quite
seamlessly as component of larger, more expressive systems. This flexibility is
what we exploit to derive \(\lambda C\). To do this, we first extend the STLC
with more expressive types through simple extentions of its grammar of simple
types.

The key difference between the STLC and \(\lambda C\) are the places in which
terms and types are allowed to be. In the STLC, terms and types are entirely
seperate entities in that terms only exist on {\em term level} and types
on {\em type level}. We return to the idea of {\em levels} later. Thus, as we
mentioned earlier, this puts a constraint on the system to only allow
\(\lambda\)-abstraction and function application to happen at term level
(viz. definition of simple type). For example, the \textit{constant function}
over types \(\mathbb{N}\) and \(\mathbb{B}\) in the STLC is defined:
\begin{center}
\begin{minipage}{0.5\textwidth}
\begin{code}
const : ℕ → 𝔹 → ℕ
const = λ n b → n
\end{code}
\end{minipage}
\end{center}
Given that {\tt const} is of type {\tt ℕ → 𝔹 → ℕ}, the term that inhabits this
type is {\tt λ n b → n}, where {\tt n} and {\tt b} are typed {\tt ℕ} and {\tt 𝔹},
respectively. Here, we are also limited by the fact that this constant function can
only be applied to natural numbers and booleans. It is not the most general
(i.e., applicable to any two terms) way to define such a function. Thus, in this
example, we can see that terms and types are treated differently and are only
related in terms of {\em inhabitance}. This relation is the reason for the
limited expressivity of the STLC and is also the method in which the STLC can be
extended into a more expressive system.

@@

Another name for \(\lambda2\) is System F \cite{}. For our purposes, we treat
\(\lambda2\) as an extension of the STLC. This system is adds {\em polymorphic}
terms to the STLC (Dependency (2)). For example, let's take the definition of
{\tt const} above and change it to define {\tt polyConst}:
\begin{center}
\begin{minipage}{0.5\textwidth}
\begin{code}
polyConst : (A B : Set) → A → B → A
polyConst A B = λ a b → a
\end{code}
\end{minipage}
\end{center}
Unlike {\tt const}, {\tt polyConst} takes four arguments, where the first two
are types. These two types (viz. {\tt A} and {\tt B}) are the corresponding
types of the third and fourth arguments of {\tt polyConst}. To clarify, we can
instantiate {\tt polyConst} with \(\mathbb{N}\) and \(\mathbb{B}\) resulting
in a function of type {\tt ℕ → 𝔹 → ℕ}, giving us:
$${\tt const} = {\tt polyConst(ℕ,𝔹)}.$$
Thus, adding polymorphism to the STLC has the effect of extending the notion
of type and term. With \(\lambda2\), we are now allowed to pass types as arguments
at term level (viz. we pass the actual types \(\mathbb{N}\) and \(\mathbb{B}\)).
It is, however, necessary to constrain this expressivity with a more expressive
type (e.g. {\tt (x : Set) → ...}). Types that appear in this fashion are
called \(\Pi\)-types. Extending the grammar for simple types with \(\Pi\)-types
results in the grammar for types ({\tt t}) in \(\lambda2\):

@@

\(\Pi\)-types are actually significantly more expressive than the ones in
\(\lambda2\). In \(\lambda2\), they are constrained to be only parameterized
over other types (e.g. {\tt Set} in {\tt ∀ (x : Set) → t}). Before we can used
\(\Pi\)-types to their full potential, we must first introduce other forms of
type and term dependency.

@@

Thus, the system we are introducing is \(\lambda\omega\), which, unlike
\(\lambda\underline{\omega}\), features a more diverse grammar of types (i.e.,
\(\lambda\underline{\omega}\) does not have \(\Pi\)-types). Given its \(\lambda2\)
component, another name for this system is System F\(\omega\). Before we introduce
the new grammar for types, we first introduce the concept of {\em levels}.

@@

We read the above example as: 42 is of type \(\mathbb{N}\) which is of type
{\tt Set} which of type {\tt Set₁} and so on. There is, however, an important
distinction for the first two levels: 42 is NOT of type {\tt Set}, while
\(\mathbb{N}\) is both of type {\tt Set}, {\tt Set₁} and onwards. Terms are
therefore treated uniquely, since they only exist as elements of types.

We can now revise the definition of types ({\tt t}) to include all instances of
{\tt Set} in our infinite universe. We notate this as {\tt Set} below:

@@

To review, if we compile all the previously mentioned systems above, we have
the system \(\lambda\omega\) (System F\(\omega\)). Our final addition, the
system of \(\lambda P\), allows us to construct {\em type families} which are
themselves types and are meant to directly correspond to propositions in
higher-order logic.

Let's take a simple example. Let's define a proposition {\tt isEven} that takes
a natural number. Another way to say this is that {\tt isEven} is a predicate
over natural numbers. When writing propositions, we write them as their equivalent
type definition (similar to {\tt List}). Consequently, the proposition
{\tt isEven} is an example of a dependent type, since its {\em meaning} depends
on the natural number passed to it (types depending on terms).

@@

This method of proof is adopted from {\em Intuitionistic Logic} \cite{}. In
this form of logic, ``truth'' corresponds to the idea of {\em type inhabitability}.
This makes sense since we use types to correspond to logical propositions
(e.g. {\tt isEven}). ``Falsity'' then naturally corresponds to the un-inhabitability
of a given type. This can be seen if we try to prove that 5 is even using
{\tt isEven}. The proof seems to get ``stuck'' at trying to prove that 1 is even.

@@

\subsection*{Extension to Homotopy Type Theory (maybe)}
\subsection*{Conclusion and Acknowledgments}

>>>>

\section{Introduction}
We present an overview of dependent types and (Martin-L\"of) type theory.
At its core, type theory seeks to serve as a possible foundation of mathematics,
one that is based on a different logical foundation. This proof-relevant
form of mathematics is the basis in which types can be used to model logical
propositions and is the core idea of type theory. As such, not only does type
theory have a viable application in mathematics but also a growing interest
in the field of computer science where types find themselves in great use in
many popular programming languages. In both fields, type theory provides a
richer world of types and expressibiilty.

The study of type theory goes hand in hand with the study of dependent types,
which is why we first take the time to familiarize ourselves with them
(Section 2). We then continue with the fundamental concepts of the type theory of
Per Martin-L\"of \cite{}. For this, we recommend the reader have a running proof
assistant, such as Agda. In our examples, we refer to Agda syntax rules, and
using a proof assistant should help develop the necessary thought processes to
successfully do type theoretic proofs.

@@

\subsection*{Simply-Typed (STLC)}
\subsubsection*{Role: Base System}
The STLC is not a dependently typed system. In fact, the term {\em dependent type}
is drawn in constrast with the term {\em simple type}, where the STLC derives
its namesake. A simple type, {\tt t}, is either a {\em primitive type}, {\tt Prim},
which is the set of all types ``pre-defined'' (e.g. lists, numbers, booleans),
or a {\em function type} between other simple types. Thus, simple types have the
following grammar:
\begin{center}
\begin{minipage}{0.5\textwidth}
\begin{code}
Prim || t → t
\end{code}
\end{minipage}
\end{center}
The terms of the STLC are the terms of the \textit{untyped \(\lambda\)-calculus},
with the exception of terms that use \textit{self-application} (i.e., Y
combinator, Omega, etc.). An example term, annotated with its type, is generally
written as follows:
\begin{center}
\begin{minipage}{0.5\textwidth}
\begin{code}
example : ℕ → ℕ
example = λ (n : ℕ) → n
\end{code}
\end{minipage}
\end{center}
Or, equivalently,
\begin{center}
\begin{minipage}{0.5\textwidth}
\begin{code}
example = λ n → n
\end{code}
\end{minipage}
\end{center}
whenever the type of the argument (viz. {\tt n}) is easily inferable. In this
case, since {\tt example} is of type {\tt ℕ → ℕ}, the argument {\tt n} has
to be of type {\tt ℕ}, therefore leaving its type annotation unnecessary.

The STLC has limited expressibilty but has the benefit of no longer having any
infinite calculations \cite{} and is thus the appropriate foundation for
deriving a type theoretic system. On top of this, the STLC lends itself quite
seamlessly as a component of larger, more expressive systems. This flexibility is
what we exploit to derive \(\lambda C\). To do this, we first extend the STLC
with more expressive types through simple extentions of its type grammar.

With this, the key difference between the STLC and \(\lambda C\) are the places
in which terms and types are allowed to be within a given expression. In the
STLC, terms and types are entirely seperate entities in that terms only exist on
{\em term level} and types on {\em type level}. We return to the idea of
{\em levels} later. Thus, as we mentioned earlier, this puts a constraint on the
system to only allow \(\lambda\)-abstraction and function application to happen
at term level (viz. definition of simple type). For example, the
{\em constant function} over types \(\mathbb{N}\) and \(\mathbb{B}\) in the STLC
is defined:
\begin{center}
\begin{minipage}{0.5\textwidth}
\begin{code}
const : ℕ → 𝔹 → ℕ
const = λ n b → n
\end{code}
\end{minipage}
\end{center}
Given that {\tt const} is of type {\tt ℕ → 𝔹 → ℕ}, the term that inhabits this
type is {\tt λ n b → n}, where {\tt n} and {\tt b} are typed {\tt ℕ} and {\tt 𝔹},
respectively. Here, we are also limited by the fact that this constant function can
only be applied to natural numbers and booleans. It is not the most general
(i.e., applicable to any two terms) way to define such a function. Thus, in this
example, we can see that terms and types are treated differently and are only
related in terms of {\em inhabitance}. This relation is the reason for the
limited expressivity of the STLC and is also the method in which the STLC can be
extended into a more expressive system.

@@

Another name for \(\lambda2\) is System F \cite{}. For our purposes, we treat
\(\lambda2\) as an extension of the STLC. This system is adds {\em polymorphic}
terms to the STLC (Dependency (2)). For example, let's take the definition of
{\tt const} above and change it to define {\tt polyConst}:
\begin{center}
\begin{minipage}{0.5\textwidth}
\begin{code}
polyConst : (A B : Set) → A → B → A
polyConst A B = λ a b → a
\end{code}
\end{minipage}
\end{center}
Unlike {\tt const}, {\tt polyConst} takes four arguments, where the first two
are types. These two types (viz. {\tt A} and {\tt B}) are the corresponding
types of the third and fourth arguments of {\tt polyConst}. To clarify, we can
instantiate {\tt polyConst} with \(\mathbb{N}\) and \(\mathbb{B}\) resulting
in a function of type {\tt ℕ → 𝔹 → ℕ}, giving us:
$${\tt polyConst(ℕ,𝔹)} = {\tt const}.$$
Thus, adding polymorphism to the STLC has the effect of extending the notion
of type and term. With \(\lambda2\), we are now allowed to pass types as arguments
at term level (viz. we pass the actual types \(\mathbb{N}\) and \(\mathbb{B}\)).
It is, however, necessary to constrain this expressivity with a more expressive
type (e.g. {\tt (x : Set) → ...}). Types that appear in this fashion are
called \(\Pi\)-types. Extending the grammar for simple types with \(\Pi\)-types
results in the type grammar of \(\lambda2\):

@@

\(\Pi\)-types are actually significantly more expressive than the ones in
\(\lambda2\). In \(\lambda2\), they are only allowed to be parameterized
over other types (e.g. {\tt Set} in {\tt ∀ (x : Set) → t}). Before we can use
\(\Pi\)-types to their full potential, we must first introduce other forms of
type and term dependency.

@@

Thus, the system we are introducing here is \(\lambda\omega\), which, unlike
\(\lambda\underline{\omega}\), features a more diverse grammar of types (i.e.,
\(\lambda\underline{\omega}\) does not have \(\Pi\)-types). Given its \(\lambda2\)
component, another name for this system is System F\(\omega\). Before we introduce
the new grammar for types, we first introduce the concept of {\em levels}.

@@

We read the above example as: 42 is of type \(\mathbb{N}\) which is of type
{\tt Set} which of type {\tt Set₁} and so on. There is, however, an important
distinction for the first two levels: 42 is NOT of type {\tt Set}, while
\(\mathbb{N}\) is of type {\tt Set}, {\tt Set₁} and so on. Terms are
therefore treated uniquely, since they only exist as elements of types.

We can now revise the definition of types to include all instances of
{\tt Set} in our infinite universe. We notate this as {\tt Set} below:

@@

To review, if we compile all the previously mentioned systems above, we have
the system \(\lambda\omega\) (System F\(\omega\)). Our final addition, the
system of \(\lambda P\), allows us to construct {\em type families} which are
themselves types and are meant to directly correspond to propositions in
higher-order logic \cite{}.

Let's take a simple example. Let's define a proposition {\tt isEven} that takes
a natural number. Another way to say this is that {\tt isEven} is a predicate
over natural numbers. When writing propositions, we write them as their equivalent
type definition (similar to how {\tt List} is defined).
%% Consequently, the proposition
%% {\tt isEven} is an example of a dependent type, since its {\em meaning} depends
%% on the natural number passed to it (types depending on terms).

@@

This method of proof is adopted from {\em Intuitionistic Logic} \cite{}. In
this form of logic, ``truth'' corresponds to the idea of {\em type inhabitability}.
This makes sense since we use types to correspond to logical propositions
(e.g. {\tt isEven}) and types are either inhabitted or un-inhabitted. Naturally,
``Falsity'' then corresponds to the un-inhabitability of a given type. This can
be seen if we try to prove that 5 is even using {\tt isEven}. The proof seems to
get ``stuck'' at trying to prove that 1 is even.

@@

\subsection*{Extension to Homotopy Type Theory (maybe)}
Earlier, we briefly discussed how types can be viewed as shapes, with their
respective terms consequently viewable as points contained with shapes. This
concept is important in understanding another way to view the concepts of type
theory: {\em homotopy type theory} (HoTT).

In fact, a lot of what HoTT seeks to do stems directly from a single concept in
regular type theory: the induction principle of identity types! This is why we
briefly discussed identity types having {\em another}, less general induction
principle. This less general induction principle is known as {\tt K} and is
defined as follows:
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
K : {A : Set} (C : {x : A} → x ≡ x → Set) →
    (∀ (x : A) → C (refl x)) →
    {x : A} (p : x ≡ x) → C p
K C c (refl x) = c x
\end{code}
\end{minipage}
\end{center}
{\tt K} is similar to {\tt J} but asserts that the only proof of equality is
{\tt refl} (i.e., the only equality is reflexive equality). In fact, we can
prove this using {\tt K}:
\begin{center}
\begin{minipage}{0.9\textwidth}
\begin{code}
UIP≡ : {A : Set} {x : A} (p : x ≡ x) → p ≡ (refl x)
UIP≡ = 
  K (λ {x} x=x → x=x ≡ (refl x))
    (λ x → refl (refl x))
\end{code}
\end{minipage}
\end{center}
As we mentioned earlier, we cannot prove UIP for identity types using {\tt J},
since the identity type passed to {\tt J} is parameterized over two possibly
different terms, while {\tt K} only applies to reflexive proofs of equality.
In some cases, {\tt K} might be all that one would need to do a given number
of proofs concerning identity types, however, it limits the overall application
and expressivity of identity types.
\subsection*{Conclusion and Acknowledgments}
